---
title: "miQC_ddqc_kidney"
output: html_document
---

Running ddqc and miQC on a paired disease-versus-healthy kidney dataset would therefore allow us to test whether these methods can distinguish true biological variation driven by disease state from low-quality cells. This is a scenario the paper does not explicitly address and one with direct clinical relevance, so this dataset helps us understand the robustness of these quality control methods.

Load in the necessary libraries
```{r include=FALSE}
library(Seurat)
library(SingleCellExperiment)
library(scater)
library(flexmix)
library(splines)
library(miQC)
```

Set the values for the healthy vs disease kidney dataset
```{r echo=TRUE}
healthy_rds <- "/projectnb/ds596/projects/Team 9/scQC_project/data/extension_kidney/GSE171314_scRNA_healthy.rds"
disease_rds <- "/projectnb/ds596/projects/Team 9/scQC_project/data/extension_kidney/GSE171314_scRNA_disease.rds"
```

Setting the parameters
```{r}
npcs_use           <- 50
k_param            <- 20
cluster_resolution <- 1.4
random_seed        <- 29
```

Load and rename cells
```{r}
healthy <- readRDS(healthy_rds)
disease <- readRDS(disease_rds)
 
healthy$condition <- "Healthy"
disease$condition <- "Disease"
 
healthy <- RenameCells(healthy, add.cell.id = "Healthy")
disease <- RenameCells(disease, add.cell.id = "Disease")
```


## Healthy Kidney Cells

Load the data
```{r}
healthy_raw <- healthy
cat("Healthy: loaded", nrow(healthy_raw), "genes and", ncol(healthy_raw), "cells\n")
```

Convert to SCE and compute QC metrics

```{r}
sce_healthy   <- as.SingleCellExperiment(healthy_raw)
mt_genes      <- grepl("^MT-", rownames(sce_healthy))
feature_ctrls <- list(mito = rownames(sce_healthy)[mt_genes])
sce_healthy   <- addPerCellQC(sce_healthy, subsets = feature_ctrls)
```

Visualize metrics before filtering

```{r}
print(plotMetrics(sce_healthy))
```

Fit mixture model

```{r}
model_healthy <- mixtureModel(sce_healthy)
```

Inspect model parameters and posterior probabilities

```{r}
print(parameters(model_healthy))
print(head(posterior(model_healthy)))
 
print(plotModel(sce_healthy, model_healthy))
print(plotFiltering(sce_healthy, model_healthy))
```

Filter cells

```{r}
sce_healthy_filt <- filterCells(sce_healthy, model_healthy,
                                keep_all_below_boundary = TRUE,
                                enforce_left_cutoff = TRUE)
message(sprintf("Healthy cells before miQC: %d", ncol(sce_healthy)))
message(sprintf("Healthy cells after miQC:  %d", ncol(sce_healthy_filt)))
```

Convert back to Seurat and reprocess

```{r}
seu_healthy_miqc <- as.Seurat(sce_healthy_filt, counts = "counts", data = NULL)
seu_healthy_miqc <- NormalizeData(seu_healthy_miqc, verbose = FALSE)
seu_healthy_miqc <- FindVariableFeatures(seu_healthy_miqc, nfeatures = 2000, verbose = FALSE)
seu_healthy_miqc <- ScaleData(seu_healthy_miqc, verbose = FALSE)
seu_healthy_miqc <- RunPCA(seu_healthy_miqc, npcs = npcs_use, verbose = FALSE)
seu_healthy_miqc <- FindNeighbors(seu_healthy_miqc, dims = 1:npcs_use,
                                  k.param = k_param, verbose = FALSE)
seu_healthy_miqc <- FindClusters(seu_healthy_miqc, resolution = cluster_resolution,
                                 algorithm = 1, random.seed = random_seed, verbose = FALSE)
seu_healthy_miqc <- RunUMAP(seu_healthy_miqc, dims = 1:20, verbose = FALSE)
```

UMAP

```{r}
print(DimPlot(seu_healthy_miqc, reduction = "umap") + ggtitle("miQC UMAP: healthy kidney"))
```

Save

```{r}
saveRDS(seu_healthy_miqc, file = "healthy_miQC_filtered.rds")
```

## Disease Kidney Cells (IgA nephropathy)

Load the data
```{r}
disease_raw <- disease
cat("Disease: loaded", nrow(disease_raw), "genes and", ncol(disease_raw), "cells\n")
```

Convert to SCE and compute QC metrics
```{r}
sce_disease   <- as.SingleCellExperiment(disease_raw)
mt_genes      <- grepl("^MT-", rownames(sce_disease))
feature_ctrls <- list(mito = rownames(sce_disease)[mt_genes])
sce_disease   <- addPerCellQC(sce_disease, subsets = feature_ctrls)
```

Visualize metrics before filtering
```{r}
print(plotMetrics(sce_disease))
```

Fit mixture model
```{r}
model_disease <- mixtureModel(sce_disease)
```

Inspect model parameters and posterior probabilities
```{r}
print(parameters(model_disease))
print(head(posterior(model_disease)))
 
print(plotModel(sce_disease, model_disease))
print(plotFiltering(sce_disease, model_disease))
```

Filter cells
```{r}
sce_disease_filt <- filterCells(sce_disease, model_disease,
                                keep_all_below_boundary = TRUE,
                                enforce_left_cutoff = TRUE)
message(sprintf("Disease cells before miQC: %d", ncol(sce_disease)))
message(sprintf("Disease cells after miQC:  %d", ncol(sce_disease_filt)))
```

Convert back to Seurat and reprocess
```{r}
seu_disease_miqc <- as.Seurat(sce_disease_filt, counts = "counts", data = NULL)
seu_disease_miqc <- NormalizeData(seu_disease_miqc, verbose = FALSE)
seu_disease_miqc <- FindVariableFeatures(seu_disease_miqc, nfeatures = 2000, verbose = FALSE)
seu_disease_miqc <- ScaleData(seu_disease_miqc, verbose = FALSE)
seu_disease_miqc <- RunPCA(seu_disease_miqc, npcs = npcs_use, verbose = FALSE)
seu_disease_miqc <- FindNeighbors(seu_disease_miqc, dims = 1:npcs_use,
                                  k.param = k_param, verbose = FALSE)
seu_disease_miqc <- FindClusters(seu_disease_miqc, resolution = cluster_resolution,
                                 algorithm = 1, random.seed = random_seed, verbose = FALSE)
seu_disease_miqc <- RunUMAP(seu_disease_miqc, dims = 1:20, verbose = FALSE)
```

UMAP
```{r}
print(DimPlot(seu_disease_miqc, reduction = "umap") + ggtitle("miQC UMAP: disease kidney"))
```

Save
```{r}
saveRDS(seu_disease_miqc, file = "disease_miQC_filtered.rds")
```

